---
title: "ST5209/X Assignment 4"
format: pdf
editor: visual
author: "Teng Siyuan"
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
```

## 1. AR polynomial

Consider the AR(2) model $$
X_t = 4 + 0.5 X_{t-1} - 0.25 X_{t-2} + W_t.
$$ {#eq-ar2}

a.  What is the autoregressive polynomial?

    $$
    \phi(z)=1-0.5z+0.25z^2
    $$

b.  What are its roots?

    Let

    $$
    \phi(z)=0
    $$

    We have

    $$
    z=1\pm\sqrt3i
    $$

c.  Is this model causal? Why?

    The model is causal since all roots are outside the unit circle.

d.  Given the representation $$
    X_t = \mu + \sum_{j=0}^\infty \psi_j W_{t-j},
    $$ solve for $\psi_0, \psi_1, \psi_2, \psi_3$. *Hint: Follow the strategy discussed in `linear_process_coefs.qmd`.*

    Using $\phi(z)\left(\sum_{j=0}^{\infty}\psi_j z^j\right)=1$

    We have

    $$
    \begin{aligned}
    &(1-0.5z+0.25z^2)(\psi_0+\psi_1z+\psi_2z^2+\psi_3z^3+...)\\
    &=\psi_0+(\psi_1-0.5\psi_0)z+(\psi_2-0.5\psi_1+0.25\psi_0)z^2+(\psi_3-0.5\psi_2+0.25\psi_1)z^3+...=1
    \end{aligned}
    $$

    Then we have

    $$
    \begin{cases}\psi_0 = 1 \\\psi_1 - 0.5\psi_0 = 0 \\\psi_2 - 0.5\psi_1 + 0.25\psi_0 = 0 \\\psi_3 - 0.5\psi_2 + 0.25\psi_1 = 0\end{cases}
    $$

    which gives

    $$
    \begin{cases}\psi_0 = 1 \\\psi_1  = 0.5 \\\psi_2 = 0 \\\psi_3 = -0.125\end{cases}
    $$

## 2. Likelihood

Consider the AR(1) model $$
X_t = 1-0.6X_{t-1} + W_t,
$$ where $W_t \sim WN(0, 0.25)$ is Gaussian white noise. We are given observations $x_1 = 0.2, x_2 = -0.3, x_3 = 0.4$.

a.  What is the mean of $X_t$?

    We have

    $$
    \mu=1-0.6\mu
    $$

    which gives

    $$
    \mu=0.625
    $$

b.  Write the conditional likelihood of this model, when conditioning on the value of $x_1$.

    We have

    $$
    X_t​∣X_{t−1}​=x_{t−1}​∼N(1−0.6x_{t−1}​, 0.25)
    $$

    $$
    \begin{aligned}
    &L(\theta\mid x_1) = \prod_{t=2}^{3} \frac{1}{\sqrt{2\pi \cdot 0.25}}\exp\left(-\frac{(x_t - (1-0.6x_{t-1}))^2}{2\times 0.25}\right) \\
    &= \frac{1}{2\pi\times 0.25}\exp\left(-\frac{(-0.3-0.88)^2 + (0.4-1.18)^2}{0.5}\right)=0.01164
    \end{aligned}
    $$

## 3. Reversibility

a.  Using `set.seed(5209)` to fix a random seed, create a sample trajectory of length $n = 500$ from the AR(2) model from Problem 1 using `arima.sim()`.

    ```{r}
    set.seed(5209)
    ar2<-arima.sim(model=list(ar=c(0.5,-0.25)),n=500,mean=4)|>
      as_tsibble()
    ar2|> autoplot(.vars = value)
    ```

b.  Reverse the time index of the vector you obtain using `rev()`.

    ```{r}
    ar2_rev<-rev(ar2$value)|>
      ts()|>
      as_tsibble()
    ar2_rev|>autoplot(.vars=value)
    ```

c.  Fit an AR(2) model to the reversed time series using `fable`. Hint: You may use the code snippet `model(AR(X ~ order(2)))`.

    ```{r}
    fit<-ar2_rev|>
      model(AR(value~order(2)))
    ```

d.  Inspect the model parameters using `tidy()`. Why are they similar to the those in @eq-ar2?

    ```{r}
    tidy(fit)
    ```

    The parameters are similar to those in Equation 1 because of the reversibility of stationary AR process.

e.  Make a forecast with $h = 10$. What does this correspond to in terms of the original time series?

    ```{r}
    fit|>
      forecast(h=10)
    ```

    This correspond to predicting the 10 values that came before $x_1$ in the original series.

## 4. Yule-Walker

a.  Write the Yule-Walker equations for the AR(2) model from Problem 1.

    $$
    \begin{cases} k=0: & \gamma(0) = 0.5\gamma(1) - 0.25\gamma(2) + \sigma^2 \\ k=1: & \gamma(1) = 0.5\gamma(0) - 0.25\gamma(1) \\ k=2: & \gamma(2) = 0.5\gamma(1) - 0.25\gamma(0) \end{cases}
    $$

b.  Arrange the equations in the following matrix form (i.e. fill in the missing entries): $$
    \left[\begin{matrix}     1 & -0.5 & 0.25  \\ -0.5 & 1.25 & 0 \\
    0.25 & -0.5 & 1
    \end{matrix}\right] \left[\begin{matrix}     \gamma(0) \\ \gamma(1) \\ \gamma(2)
    \end{matrix}\right] = \left[\begin{matrix}     \sigma^2 \\ 0 \\ 0
    \end{matrix}\right].
    $$

c.  Solve the system from part b) for $\gamma(0), \gamma(1), \gamma(2)$ numerically using `solve()`.

    ```{r}
    A <- matrix(c(
       1,    -0.5,   0.25,
      -0.5,   1.25,  0,
       0.25, -0.5,   1
    ), nrow = 3, byrow = TRUE)

    b <- c(1, 0, 0)  # sigma^2 = 1

    solve(A, b)
    ```

d.  Given $$
    \Gamma_2 = \left[\begin{matrix}     \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0)
    \end{matrix}\right],
    $$ what is the top left entry of $\Gamma_2^{-1}$ in terms of $\gamma(0)$ and $\gamma(1)$?

    We have

    $$
    \Gamma_2^{-1} = \frac{1}{\gamma(0)^2 - \gamma(1)^2}\begin{bmatrix} \gamma(0) & -\gamma(1) \\ -\gamma(1) & \gamma(0) \end{bmatrix}
    $$

    The top left entry is

    $$
    \frac{\gamma(0)}{\gamma(0)^2 - \gamma(1)^2}
    $$

e.  Write a 95% confidence interval for $\phi_1$ using your answers for d) and 3d).

    ```{r}
    phi_0<-tidy(fit)$estimate[2]
    gamma0 <- solve(A,b)[1]
    gamma1 <- solve(A,b)[2]
    sigma2<-1
    sd<- sqrt((sigma2*gamma0/(gamma0**2-gamma1**2))/500)
    q<-qnorm(0.975)
    c(phi_0-q*sd,phi_0+q*sd)
    ```

## 5. Recursive forecasting for ARMA(1, 1)

Consider the $ARMA(1,1)$ model $$
X_t - 0.5X_{t-1} = W_t + 0.5W_{t-1}.
$$ In this question, we will investigate recursive forecasting. The following code snippet generates a sequence of length $n=50$ drawn from the above model.

```{r}
set.seed(5209)
n <- 50
wn <- rnorm(n)
xt <- arima.sim(model = list(ar = 0.5, ma = 0.5), innov = wn, n = n)
```

a.  Fill in the following code snippet using equation (12.14) to generate a sequence `wn_hat`.

```{r}
wn_hat <- rep(0, n)
wn_hat[[1]] <- xt[[1]]
for (i in 2:n) {
  wn_hat[i]=xt[i]-0.5*xt[i-1]-0.5*wn_hat[i-1]
}
```

b.  Make a time plot of the log absolute difference between `wn` and `wn_hat`.

    ```{r}
    log_diff <- log(abs(wn - wn_hat))
    plot(1:n, log_diff, type = "l",
         xlab = "t", ylab = "log|wn - wn_hat|",
         main = "Log Absolute Difference between wn and wn_hat")
    ```

c.  What consequence does this have for truncated forecasts?

    The forecasts become more and more precise as t increase, suggesting ​the initialization error decays over time.

d.  Compute the truncated forecast for $X_{53}$.

    ```{r}
    X_tilde_51 <- 0.5 * xt[50] + 0.5 * wn_hat[50]
    X_tilde_52 <- 0.5 * X_tilde_51
    X_tilde_53 <- 0.5 * X_tilde_52

    cat("Truncated forecast for X_53 =", X_tilde_53, "\n")
    ```

## 6. ACF, PACF, and BLPs

Let $(X_t)$ be a mean zero stationary process with the following autocovariance values:

$$
\gamma_X(0) = 2, \gamma_X(1) = 1.4, \gamma_X(2) = 0.6, \gamma_X(3) = 0.4, \gamma_X(4) = 0.2.
$$

a.  Can $(X_t)$ be an MA(2) process? Explain why or why not.

    No. For MA(2) process, there should be cuts off to zero for ACVF after lag q. Since $\gamma_X(3) = 0.4\ne0$, it can't be MA(2) process.

b.  Can $(X_t)$ be an AR(1) process? Explain why or why not.

    No. For AR(1) process, the ACVF satisfies $\gamma(k) = \phi^k \gamma(0)$, however, we can find that $\frac{\gamma_X(4)}{\gamma_X(3)}\ne\frac{\gamma_X(3)}{\gamma_X(2)}$, which indicates this can't be AR(1) process.

c.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_3 = 2$?

    $$
    \hat{X}_4=\frac{\gamma_X(1)}{\gamma_X(0)}X_3=1.4
    $$

d.  Using the notation in part c), what is the variance of $X_4 - \hat X_4$?

    $$
    Var(X_4-\hat X_4)=Var(X_4-\frac{\gamma_X(1)}{\gamma_X(0)}X_3)=[1+(\frac{\gamma_X(1)}{\gamma_X(0)})^2]\gamma_X(0)-2\frac{\gamma_X(1)}{\gamma_X(0)}\gamma_X(1)=1.02
    $$

e.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_2 = 2$?

    $$
    \hat{X}_4=\frac{\gamma_X(2)}{\gamma_X(0)}X_2=0.6
    $$

f.  Using the notation in part e), what is the variance of $X_4 - \hat X_4$?

    $$
    Var(X_4-\hat X_4)=Var(X_4-\frac{\gamma_X(2)}{\gamma_X(0)}X_2)=[1+(\frac{\gamma_X(2)}{\gamma_X(0)})^2]\gamma_X(0)-2\frac{\gamma_X(2)}{\gamma_X(0)}\gamma_X(2)=1.82
    $$

g.  Let $\alpha_X$ denote the partial autocorrelation function of $(X_t)$. What is $\alpha_X(1)$?

    $$
    \alpha_X(1)=\frac{\gamma_X(1)}{\gamma_X(0)}=0.7
    $$

h.  What is $\alpha_X(3)$?

    We use the Yule-Walker equations because PACF at lag $h$ is by definition the last coefficient $\phi_h$ of the best linear predictor of $X_t$ given $X_{t-1}, \ldots, X_{t-h}$, and the optimality condition of this predictor directly yields the Yule-Walker system $\Gamma_h\phi = \gamma_{1:h}$

    $$
    \begin{bmatrix} \gamma(0) & \gamma(1) & \gamma(2) \\ \gamma(1) & \gamma(0) & \gamma(1) \\ \gamma(2) & \gamma(1) & \gamma(0) \end{bmatrix} \begin{bmatrix} \phi_1 \\ \phi_2 \\ \phi_3 \end{bmatrix} = \begin{bmatrix} \gamma(1) \\ \gamma(2) \\ \gamma(3) \end{bmatrix}
    $$

The PACF at lag 3 is defined as the third entry of $\Gamma_3^{-1}\gamma_{1:3}$

$$
\Gamma_3 = \begin{bmatrix} 2 & 1.4 & 0.6 \\ 1.4 & 2 & 1.4 \\ 0.6 & 1.4 & 2 \end{bmatrix}, \qquad \gamma_{1:3} = \begin{bmatrix} 1.4 \\ 0.6 \\ 0.4 \end{bmatrix}
$$

```{r}
Gamma3 <- matrix(c(2, 1.4, 0.6, 1.4, 2, 1.4, 0.6, 1.4, 2), nrow = 3, byrow = TRUE)
gamma13 <- c(1.4, 0.6, 0.4)
phi <- solve(Gamma3, gamma13)
phi[3]
```
